Cryptanalysis of \MORUS

### 2018-05-10 morusAC_00_Abstract.tex
That's it

\begin{abstract}
  \MORUS is a high-performance authenticated encryption algorithm
  submitted to the CAESAR competition, and recently selected as a
  finalist.  There are three versions of \MORUS: \MORUS[640] with
  %a 640-bit state and
  a 128-bit key, and \MORUS[1280] with
  % a 1280-bit state and
  a 128-bit or 256-bit key. For all versions the security
  claim for confidentiality matches the key size.  In this paper, we
  analyze the components of this algorithm (initialization, state
  update and tag generation), and report several results.

  As our main result, we exhibit a
  linear bias in the keystream of full \MORUS, which can be used to distinguish its output
  from random, and to recover some plaintext bits in the
  broadcast setting.  For \MORUS[1280], the bias is $2^{-76}$,
  which can be exploited after around $2^{152}$ encryptions, less than
  would be expected for a 256-bit secure cipher.  For
  \MORUS[640], the same attack results in a bias of $2^{-73}$, which does not threaten
  the security of the cipher.

% More precisely, we show that a linear correlation exists between some bits of the plaintext and ciphertext. Moreover the bias depends purely on the plaintext, and not on the secret key nor the nonce, hence restricting the single use of nonce or even rekeying does not prevent the attacks.
%
% From the nature of the linear cryptanalysis, the attack works in the known plaintext attacks. Moreover, if a fixed plaintext is encrypted multiple times, the correlation can be used to recover a 1-bit information of the plaintext. The attack principally has larger impact because it can be used to recover more unknown bits of the plaintext, provided an initial segment of the plaintext is known. 
%
  To identify this bias, we make use of rotational symmetries in \MORUS
  using linear masks that are invariant by word-rotations of the
  state.  This motivates us to introduce single-word versions of
  \MORUS called \MiniMORUS, which aid in the
  analysis. The attack has been implemented and verified on
  \MiniMORUS, where it yields a bias of $2^{-16}$.

@@

  We also study reduced versions of the initialization and
  finalization of \MORUS, aiming to evaluate the security margin of these
  components.  We show a forgery attack when finalization is
  reduced from 10 rounds to 3, and a key-recovery attack in the
  nonce-misue setting when initialization is reduced from 16 to 10
  steps.  These additional results do not threaten full \MORUS, but
  studying all aspects of the design is useful to understand its
  strengths and weaknesses.

>>>>

\begin{abstract}
  \MORUS is a high-performance authenticated encryption algorithm
  submitted to the CAESAR competition, and recently selected as a
  finalist.  There are three versions of \MORUS: \MORUS[640] with
  %a 640-bit state and
  a 128-bit key, and \MORUS[1280] with
  % a 1280-bit state and
  128-bit or 256-bit keys. For all versions the security
  claim for confidentiality matches the key size.  In this paper, we
  analyze the components of this algorithm (initialization, state
  update and tag generation), and report several results.

  As our main result, we present a
  linear correlation in the keystream of full \MORUS, which can be used to distinguish its output
  from random and to recover some plaintext bits in the
  broadcast setting.  For \MORUS[1280], the correlation is $2^{-76}$,
  which can be exploited after around $2^{152}$ encryptions, less than
  would be expected for a 256-bit secure cipher.  For
  \MORUS[640], the same attack results in a correlation of $2^{-73}$, which does not threaten
  violate the security claims of the cipher.

% More precisely, we show that a linear correlation exists between some bits of the plaintext and ciphertext. Moreover the bias depends purely on the plaintext, and not on the secret key nor the nonce, hence restricting the single use of nonce or even rekeying does not prevent the attacks.
%
% From the nature of the linear cryptanalysis, the attack works in the known plaintext attacks. Moreover, if a fixed plaintext is encrypted multiple times, the correlation can be used to recover a 1-bit information of the plaintext. The attack principally has larger impact because it can be used to recover more unknown bits of the plaintext, provided an initial segment of the plaintext is known. 
%
  To identify this correlation, we make use of rotational symmetries in \MORUS
  using linear masks that are invariant by word-rotations of the
  state.  This motivates us to introduce single-word versions of
  \MORUS called \MiniMORUS, which simplifies the
  analysis. The attack has been implemented and verified on
  \MiniMORUS, where it yields a correlation of $2^{-16}$.

@@

  We also study reduced versions of the initialization and
  finalization of \MORUS, aiming to evaluate the security margin of these
  components.  We show a forgery attack when finalization is
  reduced from 10 steps to 3, and a key-recovery attack in the
  nonce-misuse setting when initialization is reduced from 16 steps to 10.  These additional results do not threaten the full \MORUS, but
  studying all aspects of the design is useful to understand its
  strengths and weaknesses.

### 2018-05-10 morusAC_01_Introduction.tex
Some more copyediting
That's it

\subsubsection*{Related Work.}
In the \MORUS submission document, the designers discuss the security of \MORUS against
several attacks, including algebraic, differential, and guess-and-determine attacks.
The main focus is on differential properties, and not many details are given for other attack vectors.
In third-party analysis,
Mileva et~al.~\cite{balkancryptsecMilevaDV15} propose a distinguisher in the nonce-reuse setting and practically evaluate the differential behaviour of toy variants of \MORUS.
Dwivedi et~al.~\cite{cryptoeprint:2016:1053} discuss the applicability of SAT solvers for state recovery, but the resulting complexity of $2^{370}$ for \MORUS[640] is well beyond the security claim.
Dwivedi et~al.~\cite{secryptDwivediMW17} also propose key-recovery attacks for \MORUS[1280] if initialization is reduced to 3.6 out of 16 steps, and discuss the security of \MORUS against internal differentials and rotational cryptanalysis.
Salam et~al.~\cite{trustcomSalamSBDPW17} apply cube attacks to obtain distinguishers for up to 5 out of 16 steps of the initialization of \MORUS[1280] with negligible complexity.
Additionally, 
Kales et~al.~\cite{cryptoeprint:2017:1137} and
Vaudenay and Viz\'{a}r \cite{cryptoeprint:2017:1147}
independently propose state-recovery and forgery attacks on \MORUS in a nonce-misuse setting with negligible data and time complexities.
Finally, a keystream bias similar in nature to our main attack was uncovered by Minaud \cite{sacryptMinaud14} on the authenticated cipher \cipher{AEGIS} \cite{AEGIS,sacryptWuP13}, another CAESAR finalist, which shares the same overall structure as \MORUS, but uses a very different, \cipher{AES}-based state update function.

@@

In more detail, our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits across five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
As a consequence, when encrypting a fixed plaintext segment multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the bias depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \MORUS[1280] with 256-bit keys, the linear bias is $2^{-76}$
\todo{double-check bias vs. correlation everywhere}
and can be exploited using about $2^{152}$ encrypted blocks.
\todo{double-check}

@@

To obtain this result, we propose a simplified abstraction of \MORUS, called \MiniMORUS. \MiniMORUS takes advantage of certain rotational symmetries in \MORUS and simplifies the description and analysis of the attack. We then show how the attack can be extended from \MiniMORUS to the real \MORUS.
%We observe that the actual bias of the approximation is better than suggested by the construction of the characteristics.
To confirm the validity of our analysis,
we practically verified the bias of the full linear trail for \MiniMORUS, as well as the bias of trail fragments for the full \MORUS.
Our analysis is also backed by a symbolic evaluation of the full trail equation and its bias on all variants of \MORUS.

@@

\subsubsection*{Outline.}
This paper is organized as follows.
We first provide a brief description of \MORUS in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \MiniMORUS, an abstraction of \MORUS based on a certain class of rotational symmetries.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a bias in the keystream over 5 steps,
and discuss the implications of our observation for the security of \MORUS in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \Cref{sec/Conclusion}.

>>>>

\subsubsection*{Related Work.}
In the \MORUS submission document, the designers discuss the security of \MORUS against
several attacks, including algebraic, differential, and guess-and-determine attacks.
The main focus is on differential properties, and not many details are given for other attack vectors.
In third-party analysis,
Mileva et~al.~\cite{balkancryptsecMilevaDV15} propose a distinguisher in the nonce-reuse setting and practically evaluate the differential behaviour of toy variants of \MORUS.
Dwivedi et~al.~\cite{cryptoeprint:2016:1053} discuss the applicability of SAT solvers for state recovery, but the resulting complexity of $2^{370}$ for \MORUS[640] is well beyond the security claim.
Dwivedi et~al.~\cite{secryptDwivediMW17} also propose key-recovery attacks for \MORUS[1280] if initialization is reduced to 3.6 out of 16 steps, and discuss the security of \MORUS against internal differentials and rotational cryptanalysis.
Salam et~al.~\cite{trustcomSalamSBDPW17} apply cube attacks to obtain distinguishers for up to 5 out of 16 steps of the initialization of \MORUS[1280] with negligible complexity.
Additionally, 
Kales et~al.~\cite{cryptoeprint:2017:1137} and
Vaudenay and Viz\'{a}r \cite{cryptoeprint:2017:1147}
independently propose state-recovery and forgery attacks on \MORUS in a nonce-misuse setting with negligible data and time complexities.
Finally, a keystream correlation similar in nature to our main attack was uncovered by Minaud \cite{sacryptMinaud14} on the authenticated cipher \cipher{AEGIS} \cite{AEGIS,sacryptWuP13}, another CAESAR finalist, which shares the same overall structure as \MORUS, but uses a very different, \cipher{AES}-based state update function.

@@

In more detail, our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits across five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
As a consequence, when encrypting a fixed plaintext segment multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the correlation depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \MORUS[1280] with 256-bit keys, the linear correlation is $2^{-76}$
and can be exploited using about $2^{152}$ encrypted blocks.

@@

To obtain this result, we propose a simplified abstraction of \MORUS, called \MiniMORUS. \MiniMORUS takes advantage of certain rotational symmetries in \MORUS and simplifies the description and analysis of the attack. We then show how the attack can be extended from \MiniMORUS to the real \MORUS.
%We observe that the actual bias of the approximation is better than suggested by the construction of the characteristics.
To confirm the validity of our analysis,
we practically verified the correlation of the full linear trail for \MiniMORUS, as well as the correlation of trail fragments for the full \MORUS.
Our analysis is also backed by a symbolic evaluation of the full trail equation and its correlation on all variants of \MORUS.

@@

\subsubsection*{Outline.}
This paper is organized as follows.
We first provide a brief description of \MORUS in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \MiniMORUS, an abstraction of \MORUS based on a certain class of rotational symmetries.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a correlation in the keystream over 5 steps,
and discuss the implications of our observation for the security of \MORUS in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \Cref{sec/Conclusion}.

### 2018-05-10 morusAC_04_Mini_trails.tex
Some more copyediting
That's it

In the equation for \MiniMORUS[1280] from the previous section, each line on the right-hand side of the equality involves distinct $S^t_{i,j}$ terms, and each line has a weight of 1. By the piling-up lemma, it follows that if we assume distinct $S^t_{i,j}$ terms to be uniform and independent, then the expression on the right-hand side has a weight of 16. Hence the linear combination of ciphertext bits on the left-hand side has a correlation of $2^{-16}$. The same holds for \MiniMORUS[640] (cf. \Cref{sec:traileq}).

@@

In order to preserve anonymity, our implementation is attached to the submission as auxiliary material. In the full version of this paper, a link will be provided instead.

>>>>

In the equation for \MiniMORUS[1280] from the previous section, each line on the right-hand side of the equality involves distinct $S^t_{i,j}$ terms, and each line has a weight of 1. By the Piling-Up Lemma, it follows that if we assume distinct $S^t_{i,j}$ terms to be uniform and independent, then the expression on the right-hand side has a weight of 16. Hence the linear combination of ciphertext bits on the left-hand side has a correlation of $2^{-16}$. The same holds for \MiniMORUS[640] (cf. \Cref{sec:traileq}).

@@

### 2018-05-10 morusAC_05_Full_trails.tex
That's it

For \MORUS[1280], since the trail fragment $\gamma^t_i$ is used four times, this phenomenon adds a contribution of $4 \cdot 3 = 12$ to the overall weight of the full trail. This results in a total weight of $4 \cdot 16 + 12 = 76$ (recall that the weight of the trail on \MiniMORUS[1280] is 16). We have confirmed this by explicitly computing the full trail equation in \Cref{sec:traileq}, and evaluating its exact weight like we did for \MiniMORUS in \Cref{sec:minibias}. That is, since the equation is quadratic, we may view it as a graph, which we split into connected components; we then compute the weight of each connected component separately by brute force, and then add up the weights of all components per the piling-up lemma. Overall, the full trail equation given in \Cref{sec:traileq} yields a weight of 76 for the full trail on \MORUS[1280].

>>>>

For \MORUS[1280], since the trail fragment $\gamma^t_i$ is used four times, this phenomenon adds a contribution of $4 \cdot 3 = 12$ to the overall weight of the full trail. This results in a total weight of $4 \cdot 16 + 12 = 76$ (recall that the weight of the trail on \MiniMORUS[1280] is 16). We have confirmed this by explicitly computing the full trail equation in \Cref{sec:traileq}, and evaluating its exact weight like we did for \MiniMORUS in \Cref{sec:minibias}. That is, since the equation is quadratic, we may view it as a graph, which we split into connected components; we then compute the weight of each connected component separately by brute force, and then add up the weights of all components per the Piling-Up Lemma. Overall, the full trail equation given in \Cref{sec:traileq} yields a weight of 76 for the full trail on \MORUS[1280].

### 2018-05-10 morusAC_06_Discussion.tex
That's it

\subsubsection{Keystream biases.}

We emphasize that the biases we uncover between ciphertext bits are
\emph{absolute}, in the sense that they do not depend on the encryption
key, or the nonce.  This is the same situation as the keystream biases
in AEGIS~\cite{sacryptMinaud14}.  As such, they can be leveraged to
mount an attack in the broacast setting, where the same message is
encrypted multiple times with different IVs and potentially different
keys~\cite{DBLP:conf/fse/MantinS01}.  In particular, the broadcast setting appears
in practice in man-in-the-browser attacks against HTTPS connections
following the BEAST model~\cite{duong2011here}.  In this scenario, an
attacker uses Javascript code running in the victim's browser (by
tricking the victim to visit a malicious website) to generate a large
number of request to a secure website.  Because of details of the HTTP
protocol, each request includes an authentication token to identify the
user, and the attacker can target this token as a repeated plaintext.
Concretely, biases in the RC4 keystream have been exploited in this
setting, leading to the recovery of authentication cookies in
practice~\cite{DBLP:conf/uss/AlFardanBPPS13}.

@@

With the trail presented in this work, the data complexity is clearly
out of reach in practice, since exploiting the bias would require
$2^{152}$ encrypted blocks for \cipher{MORUS1280}, and $2^{146}$
encrypted blocks for \cipher{MORUS640}. The data complexity could be
slightly lowered by leveraging multilinear cryptanalysis; indeed, the
trail holds for any bit shift, and if we assume independence, we could
run $w$ copies of the trail in parallel on the same encrypted blocks
(recall that $w$ is the word size, and the trail is invariant by
rotation by $w$ bits). This would save a factor $2^5$ on the data
complexity for \cipher{MORUS640}, and $2^6$ for \cipher{MORUS1280}; but
the resulting complexity is still out of reach.

>>>>

\subsubsection{Keystream correlation.}

We emphasize that the correlation we uncover between ciphertext bits is
\emph{absolute}, in the sense that it does not depend on the encryption
key, or on the nonce.  This is the same situation as the keystream correlations
in AEGIS~\cite{sacryptMinaud14}.  As such, they can be leveraged to
mount an attack in the broacast setting, where the same message is
encrypted multiple times with different IVs and potentially different
keys~\cite{DBLP:conf/fse/MantinS01}.  In particular, the broadcast setting appears
in practice in man-in-the-browser attacks against HTTPS connections
following the BEAST model~\cite{duong2011here}.  In this scenario, an
attacker uses Javascript code running in the victim's browser (by
tricking the victim to visit a malicious website) to generate a large
number of request to a secure website.  Because of details of the HTTP
protocol, each request includes an authentication token to identify the
user, and the attacker can target this token as a repeated plaintext.
Concretely, correlations in the RC4 keystream have been exploited in this
setting, leading to the recovery of authentication cookies in
practice~\cite{DBLP:conf/uss/AlFardanBPPS13}.

@@

With the trail presented in this work, the data complexity is clearly
out of reach in practice, since exploiting the correlation would require
$2^{152}$ encrypted blocks for \cipher{MORUS1280}, and $2^{146}$
encrypted blocks for \cipher{MORUS640}. The data complexity could be
slightly lowered by leveraging multilinear cryptanalysis; indeed, the
trail holds for any bit shift, and if we assume independence, we could
run $w$ copies of the trail in parallel on the same encrypted blocks
(recall that $w$ is the word size, and the trail is invariant by
rotation by $w$ bits). This would save a factor $2^5$ on the data
complexity for \cipher{MORUS640}, and $2^6$ for \cipher{MORUS1280}; but
the resulting complexity is still out of reach.

### 2018-05-10 morusAC_07_IniFin.tex
That's it

\subsection{Forgery with Reduced Finalization}
\label{subsec/Fin}
We present forgery attacks on 3 out of 10 steps of \MORUS[1280] that claims 128-bit security for integrity. The attack only works for a limited number of steps, while it works in the nonce-respect setting. As far as we know, this is the first attempt to evaluate integrity of \MORUS in the nonce-respect setting.

\subsubsection{Overview.}
A general strategy of forgery attacks in the nonce-respect setting is to inject some difference in a message block and to propagate the difference so that it can be canceled by difference in another message block. However this approach does not work well against \MORUS because its large state size prevents the attackers from controlling the difference to be canceled by another message difference.

Here we focus on the property that the padding for an associated data $A$ and a message $M$ is the zero-padding, hence $A$ and $A'=A\|0^*$ and $M$ and $M'=M\|0$ will make the result identical for the associated data processing and the encryption parts, as long as $A,A'$ and $M,M'$ fit in the same number of blocks. During the finalization, because $A,A'$ (resp. $M,M'$) have different lengths, the corresponding $adlen$s (resp. $msglen$s) are different, which appears as $\Delta adlen$ (resp. $\Delta msglen$), and is injected from the message input. Our strategy is to propagate this difference to 128-bit tag $T$ and $T'$ such that their difference $\Delta T$ appears higher probability than $2^{-128}$. All in all, the forgery succeeds as long as the desired $\Delta T$ is obtained, in other words, the attackers do not have to cancel the state difference, which is the main advantage of attacking the finalization.

@@

\subsubsection{Differential Trails.}
Recall that the message input during the finalization of \MORUS[1280] is $adlen\|msglen\|0^{128}$ where $adlen$ and $msglen$ are 64-bit strings. We set $\Delta adlen$ to be low Hamming weight, e.g.~{\tt 0x0000000000000001}. 
%Hereafter, we use {\tt typefont} denote hexadecimal numbers.
This difference propagates during 3 steps as specified in \Cref{Tbl:fin_diff}.

Recall that each step consists of 5 rounds and the input message is absorbed to the state in rounds 2 to 5. The trail in \Cref{Tbl:fin_diff} initially does not have any difference and the same continues even after round 1. Differences start to appear from round 2 and they will go through the bitwise-AND operation from round 4. We need to pay 1-bit to control each active AND gate. The probability evaluation for round 15 can be ignored because in this round only $S_4$ is non-linearly updated, while $S_4$ is never used for computing the tag. Finally, bitwise-AND in the tag computation is taken into account. Note that the tag is only 128 LSBs, thus the number of active AND gates should be counted only in those bits. As shown in \Cref{Tbl:fin_diff}, we can have a particular tag difference $\Delta T$ with probability $2^{-88}$. Thus after observing $A$ and corresponding $T$, $A\|0$ and $(T \oplus \Delta T)$ is a valid pair with probability $2^{-88}$.

@@

\subsubsection{Remarks.} The fact that $S_4$ updated in the last round is not used to compute a tag implies that the \MORUS finalization generally includes unnecessary computations with respect to security. It may be interesting to tweak the design such that the tag can also depend on $S_4$. Indeed in \Cref{Tbl:fin_diff}, we can observe some jump-up of the probability in the tag computation. This is because the non-linearly involved terms are $S_2 \cdot S_3$, and $S_3$ that was updated 2-round before has a high Hamming weight. In this sense, involving $S_4$ in non-linear terms of the tag computation imposes more difficulties for the attacker.

\subsection{Extending State Recovery to Key Recovery}
\label{subsec/Ini}
Kales et al.~\cite{cryptoeprint:2017:1137} showed that internal state of \MORUS[640] can be recovered under the nonce-misuse scenario using $2^5$ plaintext-ciphertext pairs. As claimed by \cite{cryptoeprint:2017:1137} the attack is naturally extended to \MORUS[1280] though Kales et al. \cite{cryptoeprint:2017:1137} did not demonstrate specific attacks. The recovered state allows the attacker to mount a universal forgery attack under the same nonce. However, the key still cannot be recovered because the key is used both at the beginning and end of the finalization, which prevents the attacker to back track the state value to the initial state. In this section, we show that meet-in-the-middle attacks can allow the attacker to recover the key faster than the exhaustive search for a relatively large number of steps, i.e.~10 out of 16 steps in \MORUS[1280].

>>>>

\subsection{Forgery with Reduced Finalization}
\label{subsec/Fin}
We present forgery attacks on 3 out of 10 steps of \MORUS[1280] that claims 128-bit security for integrity. The attack only works for a limited number of steps, while it works in the nonce-respecting setting. As far as we know, this is the first attempt to evaluate integrity of \MORUS in the nonce-respecting setting.

\subsubsection{Overview.}
A general strategy for forgery attacks in the nonce-respecting setting is to inject some difference in a message block and propagate the it so that it can be canceled by a difference in another message block. However this approach does not work well against \MORUS due to its large state size which prevents an attacker from easily controlling the differences in different registers.

Here we focus on the property that the padding for an associated data $A$ and a message $M$ is the zero-padding, hence $A$ and $A'=A\|0^*$ and $M$ and $M'=M\|0$ result in identical states after the associated data processing and the encryption parts, as long as $A,A'$ and $M,M'$ fit in the same number of blocks. During the finalization, since $A,A'$ (resp. $M,M'$) have different lengths, the corresponding $adlen$s (resp. $msglen$s) are different, which appears as $\Delta adlen$ (resp. $\Delta msglen$), and is injected through the message input interface. Our strategy is to propagate this difference to the 128-bit tags $T$ and $T'$ such that their difference $\Delta T$ appears with higher probability than $2^{-128}$. All in all, the forgery succeeds as long as the desired $\Delta T$ is obtained or in other words, the attacker does not have to cancel the state difference, which is the main advantage of attacking the finalization part of the scheme.

@@

\subsubsection{Differential Trails.}
Recall that the message input during the finalization of \MORUS[1280] is $adlen\|msglen\|0^{128}$ where $adlen$ and $msglen$ are 64-bit strings. We set $\Delta adlen$ to be of low Hamming weight, e.g.~{\tt 0x0000000000000001}. 
%Hereafter, we use {\tt typefont} denote hexadecimal numbers.
This difference propagates through 3 steps as specified in \Cref{Tbl:fin_diff}.

Recall that each step consists of 5 rounds and the input message is absorbed to the state in rounds 2 to 5. The trail in \Cref{Tbl:fin_diff} initially does not have any difference and the same continues even after round 1. Differences start to appear from round 2 and they will go through the bitwise-AND operation from round 4. We need to pay 1-bit to control each active AND gate. The probability evaluation for round 15 can be ignored since in this round only $S_4$ is non-linearly updated, while $S_4$ is never used for computing the tag. Finally, bitwise-AND in the tag computation is taken into account. Note that the tag is only 128 LSBs, thus the number of active AND gates should be counted only for those bits. As shown in \Cref{Tbl:fin_diff}, we can have a particular tag difference $\Delta T$ with probability $2^{-88}$. Thus after observing $A$ and corresponding $T$, $A\|0$ and $(T \oplus \Delta T)$ is a valid pair with probability $2^{-88}$.

@@

\subsubsection{Remarks.} The fact that the $S_4$ is updated in the last round but is not used in the tag generation implies that the \MORUS finalization generally includes unnecessary computations with respect to security. It may be interesting to tweak the design such that the tag can also depend on $S_4$. Indeed in \Cref{Tbl:fin_diff}, we can observe some jump-up of the probability in the tag computation. This is because the non-linearly involved terms are $S_2 \cdot S_3$, and $S_3$ that was updated 2-round before has a high Hamming weight. In this sense, involving $S_4$ in non-linear terms of the tag computation imposes more difficulties for the attacker.

\subsection{Extending State Recovery to Key Recovery}
\label{subsec/Ini}
Kales et al.~\cite{cryptoeprint:2017:1137} showed that the internal state of \MORUS[640] can be recovered under the nonce-misuse scenario using $2^5$ plaintext-ciphertext pairs. As claimed by \cite{cryptoeprint:2017:1137} the attack is naturally extended to \MORUS[1280] though Kales et al. \cite{cryptoeprint:2017:1137} did not demonstrate specific attacks. The recovered state allows the attacker to mount a universal forgery attack under the same nonce. However, the key still cannot be recovered because the key is used both at the beginning and end of the finalization, which prevents the attacker to back track the state value to the initial state. In this section, we show that meet-in-the-middle attacks allow the attacker to recover the key faster than exhaustive search for a relatively large number of steps, i.e.~10 out of 16 steps in \MORUS[1280].

### 2018-05-10 morusAC_08_Conclusion.tex
That's it

This work provides a comprehensive analysis of the components of
\MORUS.  In particular we show that the state update has important
weaknesses with a linear bias of $2^{-76}$ between ciphertext bits and
plaintext bits in \MORUS[1280].  This enables a plaintext recovery
attack in the broadcast setting, using about $2^{152}$ blocs of data.
While the amount of data required is impractical, this seems to
violate the security claims of \MORUS[1280] because the attack works
even if the key is refreshed regularly.  Moreover, the broadcast
setting is practically relevant, as was shown with attacks aginst RC4
as used in TLS~\cite{DBLP:conf/uss/AlFardanBPPS13}.

>>>>

This work provides a comprehensive analysis of the components of
\MORUS.  In particular, we show that \MORUS[1280]'s keystream exhibits a correlation of $2^{-76}$ between certain ciphertext bits. This enables a plaintext recovery
attack in the broadcast setting, using about $2^{152}$ blocks of data.
While the amount of data required is impractical, this seems to
violate the security claims of \MORUS[1280] because the attack works
even if the key is refreshed regularly.  Moreover, the broadcast
setting is practically relevant, as was shown with attacks against RC4
as used in TLS~\cite{DBLP:conf/uss/AlFardanBPPS13}.
